---
title: "GRAMMAR"
author: "Robert Esposito"
date: "2023-10-03"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: united
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r table, include=FALSE}
library(knitr)

```


# Chapter One 

## Communication Goals
  * Ask someone's name.

  * Ask how someone is.

## Grammar Goals
  * First, second, and third person singular pronouns.

  * First, second, and third person singular present tense conjugations.

  * Ask where someone is from, say where you are from.

  * Interrogatives "imaynataq" (how) and "imataq" (what), "maymanta" (where from)

### Suffixes

```{r suffixes, include=FALSE}
library(here)
library(knitr)
suffixes_1 = read.csv(here("anki_decks","kawsay_vida","suffixes.csv"))
```

```{r suffixesprinted, include=TRUE,echo=FALSE}
knitr::kable(suffixes_1)
```

	-rí	=	and what about X?
	-pis	=	also
	-pis... -lla-taq	= and X too
	-taq	=	question tag
	-qa	=	topic marker
	-lla	=	softens/limits force

### Conjugations
	-ni	=	present tense first person singular.
	-nki	=	present tense second person singular.
	-n	=	present tense third person singular.

# CHAPTER TWO

## COMMUNICATION GOALS:
Ask who someone is.
Talk and ask about family members.

## GRAMMAR GOALS
Interrogative "pi" (who)
First, second, and third person possessive suffixes.
Present continous aspect conjugations.

### SUFFIXES
	-chu	=	absolute interrogative marker (yes/no question)
	-chu... -chu	=	either... or
	-pis... -pis	=	both... and

### CONJUGATIONS
	-chka	=	present progressive aspect conjugation

# CHAPTER THREE

## COMMUNICATION GOALS
Ask people what a certain thing is.

## GRAMMAR GOALS
Demonstrative adjectives (kay, chay, jaqay).
Form negative (mana... -chu).

### SUFFIXES
	mana... -chu	=	negation
	-taq	=	coordinating suffix, refers back to previous idea

# CHAPTER FOUR

## COMMUNICATION GOALS
Ask someone what they are doing.
Tell someone what you are doing e.g. daily tasks.

## GRAMMAR GOALS
Form subject-object-verb sentences.

### SUFFIXES
	-ta	=	object marker

# CHAPTER FIVE (A)

## COMMUNICATION GOALS
Ask someone where they come/are coming from.
Ask how/by what method they traveled from somewhere.

## GRAMMAR GOALS
First person (exclusive) plural and second person plural present tense conjugations.
Interrogative "imapi" (how/by what means of transport, lit. "in what").

### SUFFIXES
	-pi	=	"in"/"by"
			=	means of transport (e.g. autopichu jamunkichik ~ you came by car)
	-pis... taq	=	also, inclusion (e.g. payqa San Pedromanta, ñuqapis San Pedrollamantataq kani ~ she is from San Pedro, and I am from San Pedro, too)
	-taq	=	whereas, contrast (e.g. payqa La Pazmanta, ñuqataq San Pedromanta ~ she's from La Paz, whereas I'm from San Pedro)

### CONJUGATIONS
	-yku	=	present tense we (exclusive)
	-nkichik	=	present tense you (plural)


# CHAPTER FIVE (B)

## COMMUNICATION GOALS
Ask how many of something there is.

## GRAMMAR GOALS
Third person plural present tense conjugation.
Existential "kan".
Interrogative "machkha" (how many)

### SUFFIXES
	-kuna	=	plural marker
	-pi	=	"in"/"by"
			=	mode of speaking (e.g. aymarapi parlanku ~ they speak in Aymara)
			=	period of time (e.g. kimsa díapi chayamuyku ~ we [exclusive] arrive in three days)
### CONJUGATIONS
	-nku	=	present tense they