---
title: "GRAMMAR"
author: "Robert Esposito"
date: "2023-10-03"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: united
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r suffixes, include=FALSE}
library(here)
library(knitr)
library(tidyverse)
library(purrr)
library(vctrs)

all_suffixes <- read.csv(here("anki_decks","kawsay_vida","suffixes.csv"))

suffixes <- all_suffixes %>%
  filter(
    conjugation == "0"
  ) %>%
  select(
    -(conjugation:chapter_number)
  )

conjugations <- all_suffixes %>%
  filter(
    conjugation == "1"
  ) %>%
  select(
    -(conjugation:chapter_number)
  )

suffix_dfs <- list()

for(i in unique(suffixes$deck_number)) {
  output <- paste("suffixes", i, sep = ".")
  assign(output, suffixes[suffixes$deck_number==i,])
  suffix_dfs[[i]] <- output
  count <- i
}

suffix_dfs <- lapply(suffix_dfs, function(x) {
  suffixes.x <- get(x) %>% select(-deck_number)})

conjugation_dfs <- list()

for(i in unique(conjugations$deck_number)) {
  output <- paste("conjugations", i, sep = ".")
  assign(output, conjugations[conjugations$deck_number==i,])
  conjugation_dfs[[i]] <- output
}

conjugation_dfs <- list_drop_empty(conjugation_dfs)

conjugation_dfs <- lapply(conjugation_dfs, function(y) {
  conjugation.y <- get(y) %>% select(-deck_number)})
```


# Chapter One 

## Communication Goals
  * Ask someone's name.

  * Ask how someone is.

## Grammar Goals
  * First, second, and third person singular pronouns.

  * First, second, and third person singular present tense conjugations.

  * Ask where someone is from, say where you are from.

  * Interrogatives "imaynataq" (how) and "imataq" (what), "maymanta" (where from)

### Suffixes

```{r suffixes-one, include=TRUE,echo=FALSE}
knitr::kable(suffix_dfs[[1]],
             row.names = FALSE)
```

### Conjugations
```{r conjugations-one, include=TRUE,echo=FALSE}
knitr::kable(conjugation_dfs[[1]],
             row.names = FALSE)
```

# Chapter Two

## Communication Goals
  * Ask who someone is.

  * Talk and ask about family members.

## Grammar Goals
  * Interrogative "pi" (who)

  * First, second, and third person possessive suffixes.

  * Present continuous aspect conjugations.

### Suffixes
```{r suffixes-two, include=TRUE,echo=FALSE}
knitr::kable(suffix_dfs[[2]],
             row.names = FALSE)
```

### Conjugations
```{r conjugations-two, include=TRUE,echo=FALSE}
knitr::kable(conjugation_dfs[[2]],
             row.names = FALSE)
```

# Chapter Three

## Communication Goals
  * Ask people what a certain thing is.

## Grammar
  * Demonstrative adjectives (kay, chay, jaqay).

  * Form negative (mana... -chu).

### Suffixes
```{r suffixes-three, include=TRUE,echo=FALSE}
knitr::kable(suffix_dfs[[3]],
             row.names = FALSE)
```

# Chapter Four

## Communication Goals
  * Ask someone what they are doing.

  * Tell someone what you are doing e.g. daily tasks.

## Grammar Goals
  * Form subject-object-verb sentences.

### Suffixes
```{r suffixes-four, include=TRUE,echo=FALSE}
knitr::kable(suffix_dfs[[4]],
             row.names = FALSE)
```

# Chapter Five (A)

## Communication Goals
  * Ask someone where they come/are coming from.

  * Ask how/by what method they traveled from somewhere.

## Grammar Goals
  * First person (exclusive) plural and second person plural present tense conjugations.

  * Interrogative "imapi" (how/by what means of transport, lit. "in what").

### Suffixes
```{r suffixes-five, include=TRUE,echo=FALSE}
knitr::kable(suffix_dfs[[5]],
             row.names = FALSE)
```

### Conjugations
```{r conjugations-three, include=TRUE,echo=FALSE}
knitr::kable(conjugation_dfs[[3]],
             row.names = FALSE)
```

# Chapter Five (B)

## Communication Goals
  * Ask how many of something there is.

## Grammar Goals
  * Third person plural present tense conjugation.

  * Existential "kan".

  * Interrogative "machkha" (how many)

### Suffixes
```{r suffixes-six, include=TRUE,echo=FALSE}
knitr::kable(suffix_dfs[[6]],
             row.names = FALSE)
```

### Conjugations
```{r conjugations-four, include=TRUE,echo=FALSE}
knitr::kable(conjugation_dfs[[4]],
             row.names = FALSE)
```

# Chapter Five (B)

## Communication Goals
  * Talk about where you and people are going
  * Talk about going somewhere and doing something

## Grammar Goals
  * Interrogative "maymantaq" (where, as in, "where are you going?)

### Suffixes
```{r suffixes-seven, include=TRUE,echo=FALSE}
knitr::kable(suffix_dfs[[7]],
             row.names = FALSE)
```

# Chapter Six (A)

## Communication Goals
  * 

## Grammar Goals
  * 

### Suffixes